---
title: "Final assignment"
author: "Stefano Galeano"
date: "29 maggio 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Context and questions

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

“Is an automatic or manual transmission better for MPG”
"Quantify the MPG difference between automatic and manual transmissions"

### Data Overview

The dataset is composed of 11 variables for each observation:

* **mpg** fuel consumption [Miles/(US) gallon]
* **cyl** number of cylinders (2,4 and 8)
* **disp** Displacement (cu.in.)
* **hp** wross horsepower
* **drat** Rear axle ratio
* **wt** weight [1000 lbs]
* **qsec** 1/4 mile time
* **vs**  engine type (0 = V, 1 = straight)
* **am** transmission type (0 = automatic, 1 = manual)
* **gear** number of forward gears
* **carb** number of carburetors

The first few rows of the dataset are shown below:

```{r}
head(mtcars)

```
